Phenotypic evolution is more constrained in simpler food webs

### 2019-01-11 manuscript.Rmd
remove old draft of text from Abstract and Intro

Global change is simplifying the structure of ecological networks; however, we are currently in a poor position to predict how these simplified communities will affect the evolutionary potential of remaining populations. Theory on adaptive landscapes provides a framework for predicting how selection constrains phenotypic evolution, but often treats the community context of evolving populations as a ``black box''. <!--In contrast, ecological networks provide an explicit representation of the community context, yet often ignore the role of natural selection.-->Here, we integrate ecological networks and adaptive landscapes to examine how changes in food-web complexity shape evolutionary constraints. We conducted a field experiment that manipulated the diversity of insect parasitoids (food-web complexity) that were able to impose selection on an insect herbivore. We then measured herbivore survival as a function of three key phenotypic traits. We found that more traits were under selection in simpler vs. more complex food webs. The adaptive landscape was more neutral in complex food webs because different parasitoid species impose different selection pressures, minimizing relative fitness differences among phenotypes. Our results suggest that phenotypic evolution becomes more constrained in simplified food webs. This indicates that the simplification of ecological communities may constrain the adaptive potential of remaining populations to future environmental change.

@@

The adaptive landscape provides a powerful framework for understanding how natural selection has shaped the evolution of biodiversity ---from genes to phenotypes to species [@Wright1931; @Simpson1944; @Arnold2001]. More than a metaphor, the adaptive landscape links quantitative genetic and phenotypic variation to evolution by natural selection [@Lande1979; @Arnold1984applications; @Arnold1984theory]. Ecological interactions often play a key role in shaping adaptive landscapes, as evidenced by the role of antagonistic [@Schluter2000; @Abrams2000] and mutualistic [@Bronstein2006] interactions in driving evolutionary change.<!--, often play a key role in shaping the adaptive landscape and phenotypic evolution in multiple traits [; ;]. For example, competition for shared prey in threespine stickleback has driven their adaptive divergence into ecotypes with specialized body and gill morphologies [@Schluter2000, other papers]. Similarly, the elongated, red corollas of many flowering species have evolved in response to the efficient pollination of hummingbirds [cite].--> Although there is clear evidence that pairwise interactions can shape the adaptive landscape, we do not know how the adaptive landscape is shaped by the community context [@McPeek2017; @terHorst2018]<!--SOFTEN THIS STATEMENT-->. Resolution on how the community context shapes phenotypic evolution is urgently needed though, given the rapid impacts of climate change on ecological communities [@Scheffers2016].

Ecological networks, such as food webs describing who-eats-whom in ecological communities, provide an explicit representation of the community context [@Bascompte2014; @McCann2012]<!--SENTENCE SEEMS MISPLACED-->. <!--These networks describe the interdependency of populations within an ecological community, providing an effective framework for predicting how the strength and organization of ecological interactions (network structure) affect population and community dynamics [@Bascompte2014; @McCann2012]. Nevertheless, these networks often lack quantitative information on how ecological interactions influence selection on phenotypic traits. Thus, it remains unclear how network structure shapes adaptive landscapes of interacting populations. This represents a major barrier in our current ability to predict how changes in community context will affect phenotypic evolution.--> Here, we integrate ecological networks and adaptive landscapes to understand how the community context constrains evolutionary change [@Arnold1992]. Different aspects of evolutionary constraints can be inferred by quantifying the slope and curvature of the adaptive landscape. For example, the slope is determined by directional selection gradients acting on each phenotypic trait and influences the trajectory of evolutionary change [@Lande1979;@Arnold1992]. Evolutionary trajectories become more constrained with an increase in the number of traits under selection, as this diminishes the number of optimal solutions [@Arnold2003]. The curvature of the adaptive landscape can also constrain evolution through its indirect effect on genetic constraints [@Arnold1992; @Hansen2008]. Genetic constraints are largely governed by a population's **G**-matrix ---the additive genetic variances and covariances between traits [@Hansen2008]. In general, genetic constraints will increase with the number of traits under directional or stabilizing selection, as this will decrease the additive genetic variance in those traits [@Hansen2008]. Genetic constraints may also increase with the number of trait combinations under correlational selection, as this type of selection decreases the evolutionary independence of traits [@Hansen2008]. If we want to predict how the community context constrains evolutionary change, we must understand how ecological networks shape the adaptive landscape.
<!-- need to cite more than just Hansen and Houle 2008 here -->

Global change is simplifying the structure of ecological communities, which may influence evolutionary constraints in a number of ways. For example, in a multitrophic community or food web, if different consumers impose directional selection on different traits of a shared resource, then more complex food webs may constrain evolution by increasing the number of traits under selection. Alternatively, if consumers impose selection on different values of a trait, then their selective effects would cancel each other out in more complex food webs. To examine these different possibilities, <!--To examine the effect of network structure on adaptive landscapes, -->we conducted a field experiment that manipulated the complexity of trophic interactions (hereafter food-web complexity) associated with an abundant insect herbivore (*Iteomyia salicisverruca*, Family Cecidomyiidae; fig. \ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped galls when they feed on the developing leaves of willow trees [*Salix* sp., @Russo2006]. These galls protect larva from attack by generalist predators (e.g. ants, spiders), but they suffer high mortality from egg and larval parasitoids [@Barbour2016]. We manipulated food-web complexity by either excluding the guild of larval parasitoids (simple food web) or allowing both egg and larval parasitoids to impose selection on gall midge traits (complex food web; fig. \ref{fig:Conceptual}). <!--Food-web complexity may influence the adaptive landscape in a number of ways. For example, if each parasitoid imposes directional selection on a different trait, then more complex food webs may constrain evolution by increasing the number of traits under selection. Alternatively, if parasitoids impose selection on different values of a trait, then their selective effects would cancel each other out in more complex food webs. To examine these different possibilities (among others), -->We then applied modern statistical methods to quantify the effect of food-web complexity on the slope and curvature of the gall midge's adaptive landscape. Taken together, our study gives insight to how local extinctions, and concomitant decreases in food-web complexity, may constrain the evolution of interacting populations.

@@

<!--
Here, we integrate ecological networks and the adaptive landscape to understand how changes in network structure affect the dynamics of natural selection. In particular, we are interested in how network structure constrains or enables phenotypic evolution [@Arnold1992]. We can gain insight to the selective constraints imposed by network structure by examining its effect on the slope and curvature of the adaptive landscape. To illustrate this, let us examine the adaptive landscape of two hypothetical traits, $z_i$ and $z_j$. The slope of the adaptive landscape is described by a column vector of directional selection gradients acting on these traits (cite):

$$\text{Slope} = \beta = \begin{pmatrix} \beta_{z_i} \\ \beta_{z_j} \end{pmatrix} $$

If each phenotypic trait experiences directional selection ($\beta_{z_i},\beta_{z_j}\neq0$), then this implies that there is an optimal phenotype (adaptive peak) and thus strong constraints on the trajectory of evolution. In contrast, if selection is absent on a trait (say $\beta_{z_i}=0$), then there may be multiple peaks (a line across all possible $z_i$ in this example) where fitness is equivalent and thus weaker selective constraints [@Arnold1992]. The curvature of the adaptive landscape may also constrain phenotypic evolution through its indirect effect on genetic constraints [@Arnold1992]. Genetic constraints on phenotypic evolution are largely governed by a population's **G**-matrix ---the additive genetic variances and covariances between traits [@Hansen2008]. For our hypothetical example, the **G**-matrix takes the form:

$$\text{G} = \begin{pmatrix} V_{z_i:z_i} & \\ V_{z_i:z_j} & V_{z_j:z_j} \end{pmatrix} $$

where $V_{z_i:z_i}$ is the additive genetic variance in $z_i$ and $V_{z_i:z_j}$ is the additive genetic covariance between $z_i$ and $z_j$. Changes in the **G**-matrix within a generation are governed by the following equation [cite]:

$$\Delta \text{G}=\text{G}(\gamma-\beta \beta^{\text{T}})\text{G}$$
where $\gamma$ corresponds to the matrix of quadratic ($\gamma_{z_i:z_i}$) and correlational ($\gamma_{z_i:z_j}$) selection gradients. The terms in parentheses in the above equation describe the curvature (C) of the adaptive landscape, or the effects of nonlinear selection. To make explicit how landscape curvature affects the **G**-matrix, we rewrite the curvature matrix as:

$$\text{C}= \begin{pmatrix} \gamma_{z_i:z_i} & \\ \gamma_{z_i:z_j} & \gamma_{z_j:z_j} \end{pmatrix} - \begin{pmatrix} \beta_{z_i}\beta_{z_i} & \\ \beta_{z_i}\beta_{z_j} & \beta_{z_j}\beta_{z_j} \end{pmatrix} $$
$$\text{C} = \begin{pmatrix} C_{z_i:z_i} & \\ C_{z_i:z_j} & C_{z_j:z_j} \end{pmatrix}$$
The diagonal terms of this curvature matrix describe how selection changes the additive genetic variance of traits ($V_{z_i:z_i}$). If selection acts to decrease the additive genetic variance, which is common when there is no disruptive selection ($\gamma_{z_i:z_i}>0$), then this will constrain the trait's evolutionary response to future selection [@Hansen2008]. The off-diagonal terms of the curvature matrix are also important as they describe the change in additive genetic covariances between traits. Nonzero values along the off-diagonal indicate that selection favors trait integration, which may result in stronger genetic constraints [@Hansen2008].

To examine how network structure shapes the adaptive landscape, we conducted a field experiment that manipulated the complexity of a food web associated with an abundant leaf-galling midge (*Iteomyia salicisverruca*; Family Cecidomyiidae)(fig. \ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped galls when they feed on the developing leaves of willow trees [*Salix* sp., @Russo2006]. These galls protect larva from attack by generalist predators (e.g. ants, spiders), but they suffer a high degree of mortaltiy from four parasitoid wasps. To manipulate food-web complexity, we experimentally excluded the diversity of parasitoids that were able to impose selection on gall midge larvae (fig. \ref{fig:Conceptual}).

Food-web complexity may shape the adaptive landscape, and thus selective constraints, in a number of ways. For example, if each consumer imposes directional selection on a different trait, then more complex food webs may increase the number of traits under selection, thereby increasing selective constraints. Alternatively, if consumers impose selection on different values of a trait, then their selective effects may cancel each other in more complex food webs, which would alleviate selective constraints. Food-web complexity may also indirectly affect genetic constraints. For example, if each consumer imposes selection on a different rare phenotype (i.e. each end of a trait distribution), then there would be stabilizing selection for an optimal phenotype, which would increase genetic constraints. Alternatively, if consumers are adapted to attacking the population's mean phenotype, then the relative fitness of rarer phenotypes would increase (disruptive selection), which would decrease genetic constraints. To tease apart these diverse possibilities, we applied modern statistical methods to quantify the effect of food-web complexity on the slope and curvature of the gall midge's adaptive landscape. Taken together, our study gives insight to how local extinctions, and concomitant decreases in food-web complexity, shape the adaptive landscape of interconnected populations.
-->

<!-- Old last paragraph
Here, we integrate ecological networks and the adaptive landscape to understand how changes in network structure affect the dynamics of natural selection. We conducted a field experiment that manipulated the diversity of insect parasitoids that were able to impose selection on an abundant leaf-galling midge (*Iteomyia salicisverruca*; Family Cecidomyiidae)(fig. \ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped galls when they feed on the developing leaves of willow trees [*Salix* sp., @Russo2006]. Since these galls provide protection from generalist predators (e.g. ants, spiders), our experiment provides an effective manipulation of the diversity of trophic interactions, or food-web complexity, that this insect herbivore experiences. We then applied modern statistical methods to test how food-web complexity shapes the slope and curvature of the herbivore's adaptive landscape. Taken together, our study gives insight to how local extinctions, and concomitant decreases in food-web complexity, shape the adaptive landscape of interconnected populations.
-->

<!-- Useful paragraph explaining the effect of increased mortality on strength of selection
Also, unless consumers are equivalent, then we may generally expect lower average survival in complex food webs, which would increase the strength of selection. The logic behind this is made clear if we rewrite the selection gradient, which is typically expressed as the covariance between relative fitness ($w$) and phenotype ($z_i$), $\beta_{z_i}=\textrm{cov}(w,z_i)$, as a function of both absolute fitness ($W$) and mean fitness ($\bar W$), $\beta_{z_i}=\textrm{cov}(W,z_i)/ \bar W$ [@Hunter2018]. This alternative expression makes clear that lower survival, and thus mean fitness, increases the strength of selection. Note that if consumers are equivalent in their selective effects (i.e. functionally redundant), then we would not expect any effect of food-web complexity on the adaptive landscape.
-->

>>>>

Global change is simplifying the structure of ecological networks; however, we are currently in a poor position to predict how these simplified communities will affect the evolutionary potential of remaining populations. Theory on adaptive landscapes provides a framework for predicting how selection constrains phenotypic evolution, but often treats the community context of evolving populations as a ``black box''. Here, we integrate ecological networks and adaptive landscapes to examine how changes in food-web complexity shape evolutionary constraints. We conducted a field experiment that manipulated the diversity of insect parasitoids (food-web complexity) that were able to impose selection on an insect herbivore. We then measured herbivore survival as a function of three key phenotypic traits. We found that more traits were under selection in simpler vs. more complex food webs. The adaptive landscape was more neutral in complex food webs because different parasitoid species impose different selection pressures, minimizing relative fitness differences among phenotypes. Our results suggest that phenotypic evolution becomes more constrained in simplified food webs. This indicates that the simplification of ecological communities may constrain the adaptive potential of remaining populations to future environmental change.

@@

The adaptive landscape provides a powerful framework for understanding how natural selection has shaped the evolution of biodiversity ---from genes to phenotypes to species [@Wright1931; @Simpson1944; @Arnold2001]. More than a metaphor, the adaptive landscape links quantitative genetic and phenotypic variation to evolution by natural selection [@Lande1979; @Arnold1984applications; @Arnold1984theory]. Ecological interactions often play a key role in shaping adaptive landscapes, as evidenced by the role of antagonistic [@Schluter2000; @Abrams2000] and mutualistic [@Bronstein2006] interactions in driving evolutionary change. Although there is clear evidence that pairwise interactions can shape the adaptive landscape, we do not know how the adaptive landscape is shaped by the community context [@McPeek2017; @terHorst2018]<!--SOFTEN THIS STATEMENT-->. Resolution on how the community context shapes phenotypic evolution is urgently needed though, given the rapid impacts of climate change on ecological communities [@Scheffers2016].

Ecological networks, such as food webs describing who-eats-whom in ecological communities, provide an explicit representation of the community context [@Bascompte2014; @McCann2012]. Here, we integrate ecological networks and adaptive landscapes to understand how the community context constrains evolutionary change [@Arnold1992]. Different aspects of evolutionary constraints can be inferred by quantifying the slope and curvature of the adaptive landscape. For example, the slope is determined by directional selection gradients acting on each phenotypic trait and influences the trajectory of evolutionary change [@Lande1979;@Arnold1992]. Evolutionary trajectories become more constrained with an increase in the number of traits under selection, as this diminishes the number of optimal solutions [@Arnold2003]. The curvature of the adaptive landscape can also constrain evolution through its indirect effect on genetic constraints [@Arnold1992; @Hansen2008]. Genetic constraints are largely governed by a population's **G**-matrix ---the additive genetic variances and covariances between traits [@Hansen2008]. In general, genetic constraints will increase with the number of traits under directional or stabilizing selection, as this will decrease the additive genetic variance in those traits [@Hansen2008]. Genetic constraints may also increase with the number of trait combinations under correlational selection, as this type of selection decreases the evolutionary independence of traits [@Hansen2008]. If we want to predict how the community context constrains evolutionary change, we must understand how ecological networks shape the adaptive landscape.

Global change is simplifying the structure of ecological communities, which may influence evolutionary constraints in a number of ways. For example, in a multitrophic community or food web, if different consumers impose directional selection on different traits of a shared resource, then more complex food webs may constrain evolution by increasing the number of traits under selection. Alternatively, if consumers impose selection on different values of a trait, then their selective effects would cancel each other out in more complex food webs. To examine these different possibilities, we conducted a field experiment that manipulated the complexity of trophic interactions (hereafter food-web complexity) associated with an abundant insect herbivore (*Iteomyia salicisverruca*, Family Cecidomyiidae; fig. \ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped galls when they feed on the developing leaves of willow trees [*Salix* sp., @Russo2006]. These galls protect larva from attack by generalist predators (e.g. ants, spiders), but they suffer high mortality from egg and larval parasitoids [@Barbour2016]. We manipulated food-web complexity by either excluding the guild of larval parasitoids (simple food web) or allowing both egg and larval parasitoids to impose selection on gall midge traits (complex food web; fig. \ref{fig:Conceptual}). We then applied modern statistical methods to quantify the effect of food-web complexity on the slope and curvature of the gall midge's adaptive landscape. Taken together, our study gives insight to how local extinctions, and concomitant decreases in food-web complexity, may constrain the evolution of interacting populations.

@@
